Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change

### 2025-02-06 Manuscript/endo_herbarium_R1.tex
finished edits and comments

\documentclass[11pt]{article}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,round,sectionbib,sort]{natbib}   % omit 'round' option if you prefer square brackets
\let\cite\citep

@@

	\noindent{\footnotesize Prepared using the suggested \LaTeX{} template for \textit{Am.\ Nat.}}

@@

	\section*{Abstract}
\linelabel{R2-C15-begin}Species' distributions and abundances are shifting in response to ongoing global climate change. 
Mutualistic microbial symbionts can provide hosts with protection from environmental stress that may contribute towards \revise{resilience under} environmental change, however \revise{this change} may also disrupt species interactions and lead to declines in hosts and/or symbionts.
Symbionts preserved within natural history specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes) have changed over time in response to climate change, and how these changes vary across host species' \revise{distributions}.
Specifically, we \revise{examined} 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, \emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
\revise{Analysis of an \jacob{approximate Bayesian spatially-varying coefficients model implemented in INLA}{This is the first time INLA is mentioned. It seems to be a suggestion from the reviewer. However, I think most readers may not be familiar with it. I appreciate the precision regarding the type of Bayesian statistics (approximate), but I wonder if that might be too technical for the abstract. I would understand if it were included in the methods section.} revealed} that endophytes increased in prevalence over the last two centuries from ca. 25\% to ca. 75\% prevalence, on average, across three host species.
Changes in seasonal climate drivers were associated with increasing endophyte prevalence. 
Notably, increasing precipitation during the peak growing season for \emph{Agrostis} species and decreasing precipitation for \emph{E. virginicus} were associated with increasing endophyte prevalence.
Changes in the variability of precipitation and temperature during off-peak seasons were also important predictors of increasing endophyte prevalence. 
Our analysis performed favorably in an out-of-sample predictive test with contemporary survey data, a rare extra step in collections-based research.
However, we identified greater local-scale variability in endophyte prevalence in contemporary data compared to model predictions based on historic data, suggesting new directions that could improve predictive accuracy.
Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through \revise{fitness benefits to symbiotic hosts}. \linelabel{R2-C15-end}

@@

Natural history specimens, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial and temporal scales \citep{meineke2018unrealized}. 
Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines, primarily comprised of physical specimens of organisms along with information about the time and place of their collection. 
These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \linelabel{R2C19-begin}\revise{by examining reproductive structures} \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \revise{through examination of pollen removal} \citep{pauw2011reconstruction, duan2019century}, and herbivory \revise{by documenting leaf damage} \citep{meineke2019herbarium} related to anthropogenic climate change. \linelabel{R2C19-end}
However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: microbial symbiosis.

Microbial symbionts are common to all macroscopic organisms and can have important effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts that can ameliorate environmental stress. 
For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading the interaction to deteriorate \citep{aslan2013mutualism, fowler2024microbial}. 
Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the composition of their algal symbiont community \citep{meyer2022climate}.
How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions \citep{frederickson2017mutualisms}.
Previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualist partners \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.

Understanding of how microbial symbioses are affected by climate change is additionally complicated by spatial heterogeneity in the direction and magnitude of environmental change \cite{ipcc_2021}. 
Beneficial symbionts are likely able to shield their hosts from environmental stress in locations that experience a small degree of change, but symbionts in locations that experience changes of large magnitude may be pushed beyond their physiological limits \cite{webster2008temperature}.
Additionally, symbionts are often unevenly distributed across their hosts' distribution.
Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across their environments \cite{frade2008variation, rolshausen2018quantifying, fowler2023geographic}.
Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions for where to expect changes in the symbiosis in future climates.

\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grass species \citep{leuchtmann1992systematics}.
They are transmitted vertically from maternal plants to offspring through seeds.
Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
Over time, endophytes that act as mutualists should rise in prevalence within a host population \citep{donald2021context}. 
\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
Previous survey studies of contemporary populations have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
We predicted that prevalence should track temporal changes in environmental drivers \linelabel{R2C21-begin}\revise{(i.e. drought)} \linelabel{R2C21-end} that elicit strong fitness benefits.
%For example, endophyte-mediated drought tolerance should lead endophyte prevalence to increase in regions where precipitation has declined over time.

Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of host species that harbor these symbionts \citep{white1985endophyte}, establishing that endophyte symbiosis could be identified in plant tissue from as early as 1851.
However, no subsequent studies, to our knowledge, have used the vast resources of biological collections to quantitatively assess spatio-temporal trends in endophyte prevalence and their environmental correlates. 
Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets. 
This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change. 
%Research using historical collections has clearly demonstrated other ecological signatures of a changing climate. 
However, the predictive ability derived from historical analyses is rarely tested against contemporary data \citep{lee2024phenological}. 
Critically evaluating whether insights from historical reconstruction are predictive of variation across contemporary populations is a crucial step for the field to move from reading signatures of the past to forecasting ecological dynamics into the future.

In this study, we assessed the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three North American host grass species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}).
We first addressed questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over the past two centuries? and (ii) How spatially variable are temporal trends in endophyte prevalence across eastern North America?
We then addressed how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between temporal trends in endophyte prevalence and associated changes in climate drivers?
\linelabel{R2C21-b-begin}We predicted that aggregate endophyte prevalence would increase over time in tandem with climate warming, and that hotspots of endophyte change would correspond spatially to hotspots of climate change. \linelabel{R2C21-b-end}
Finally, we evaluated \revise{(iv) how our model, built on data from historic specimens, performed in an out-of-sample test - }using data on endophyte prevalence from contemporary surveys of host populations. 
To answer these questions we examined a total of $2,346$ historic specimens collected across eastern North America between 1824 and 2019, and evaluated model performance against contemporary surveys comprising 1,442 individuals from 63 populations collected between 2013 and 2020.

\section*{Methods}
        \subsection*{Focal species}
Our surveys focused on three native North American grasses: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus} that host \emph{Epichloë} symbionts. 
\linelabel{R2C23-begin}\revise{These cool-season grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States (Fig. \ref{fig:map}). Cool-season grasses typically grow actively during the cooler temperatures of spring and autumn due to their reliance on $C_3$ photosynthesis.}\linelabel{R2C23-end}
\emph{A. hyemalis} is a small short-lived perennial species that germinates in spring and typically flowers between March and July (most common collection month: May).
\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late summer and early autumn (most common collection month: September). 
\emph{A. perennans} is more sparsely distributed, tending to be found in shadier and more moist habitats, while \emph{A. hyemalis} is commonly found in open and recently disturbed ground. 
Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A. perennans} has a slighty more northern distribution, whereas \emph{A. hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed than the \emph{Agrostis} species. 
It begins flowering as early as March or April but continues throughout the summer (most common collection month: July).

\linelabel{R1C3-begin}
\revise{Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
The fungal symbionts primarily reproduce asexually and are passed from mother to offspring by vertical transmission through seeds. 
These traits contribute to highly specialized interactions between symbiont and host.
Some host species have been shown to partner with multiple symbiont species in these genus, and in some cases multiple symbiont species can co-exist within a host population \cite{mc2014species}.
However, suveys have typically found limited \emph{Epichloë} genotypic diversity within host populations\cite{treindl2023genetic}. 
Across host populations, concentrations of biologically-active alkaloids and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.
In this analysis, we focus on the presence/absence of \emph{Epichloë} symbionts, and we discuss potential implications of hidden genotypic diversity in the Discussion.}
\linelabel{R1C3-end}

		\subsection*{Herbarium surveys}
We visited nine herbaria between 2019 and 2022 (see Table \ref{table:herbaria} for a summary of specimens included from each collection). 
With permission from herbarium staff, we acquired seed samples from $1135$ \emph{A. hyemalis} specimens collected between 1824 and 2019, $357$ \emph{A. perennans} specimens collected between 1863 and 2017, and $854$ \emph{E. virginicus} specimens collected between 1839 and 2019 (Fig. \ref{fig:map}, Fig. \ref{fig:temporal}A, Fig. \ref{fig:endo_status_map}).
%Our sampling plan was designed to minimize damage to these specimens.
We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies. 
We collected up to 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
We excluded specimens for which information about the collection location and date were unavailable.
Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
Many specimens had digitized collection information readily available, but for those that did not, we transcribed information printed on the herbarium sheet. 
Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
For fifteen of the oldest specimens, only information at the state level was available, and so we used the state centroid.
\linelabel{R2C24-begin}\revise{The median pairwise distance between georeferenced coordinate points was 841 km.
The median longitudinal width of the bounding boxes generated to geocode municipality, county, or state centroids was 44.7 km. 
Among those specimens geo-referenced at the state level, the largest bounding box, spanning the state of Texas, was 1233 km wide.
The smallest bounding boxes were less than 1 km across for small municipalities (while this suggests high precision, we note that some specimens were collected in natural habitat nearby to small munipalities not encompassed by this bounding boxes).}\linelabel{R2C24-end}

\linelabel{R1C11-begin}
\revise{Our visits focused on herbaria with historic strengths in \emph{Poaceae} collections (e.g. Texas A\&M, Missouri Botanic Garden) and other herbaria in the Southern Great Plains region of the United States. 
While these nine herbaria garnered specimens that span the focal species' ranges, our dataset unevenly samples across the study region. 
Texas, Oklahoma, Louisiana, and Missouri are the most represented states.
Uneven sampling is most likely to be consequential for \emph{A. perennans}, which has much of its range in the northeastern US. 
We explore the potential influence of spatial bias in sampling on our results through a simulation analysis (Appendix A - Supporting Methods).}\linelabel{R1C11-end}

After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae in each specimen using microscopy. 
We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue-lactic acid stain and squashed them under a microscope cover slip. 
We examined the squashed seeds for the presence of fungal hyphae at 200-400X magnification \cite{bacon2018stains}.
%In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded for that specimen.
On average we scored $4.7$ intact seeds per specimen of \emph{A. hyemalis}, $4.2$ seeds per specimen of \emph{A. perennans}, and $3.8$ seeds per specimen of \emph{E. virginicus}; we scored $10,342$ seeds in total.
\linelabel{R2C25-begin}Due to imperfect vertical transmission, \revise{the production of symbiont-free offspring from symbiotic hosts} \cite{afkhami2008symbiosis},\linelabel{R2C25-end} it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if \emph{Epichloë} hyphae were observed in none of its seeds. 
\linelabel{R2C26-begin}
To capture uncertainty in the endophyte \revise{identification} process, we recorded both a "liberal" and a "conservative" endophyte score for each plant specimen.  
\revise{When we confidently identified endophytes within a specimen's seeds, we assigned matching liberal and conservative scores.}
When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded \revise{ a positive identification} for the liberal score
and a negative \revise{identification for} the conservative score.
\revise{The liberal status assumed a potential endophyte identification was more likely to be endophyte-positive while the conservative status assumed that the potential endophyte identification was less likely to be endophyte-positive}. 
$89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
The following analyses used the liberal status, \revise{however} we repeated all analyses with the conservative status which yielded qualitatively similar results (Fig. \ref{fig:conservative_scores_plot}).\linelabel{R2C26-end}

@@

\linelabel{R2C40-begin}Our examination of historic plant specimens revealed \revise{previously hidden} \linelabel{R2C40-end}shifts in microbial symbiosis over the last two centuries. 
For the three host species we examined, there have been strong increases in prevalence of\linelabel{R2C41-begin} \revise{\emph{Epichloë} endophyte symbiosis}.\linelabel{R2C41-end}
We interpret increases in prevalence of \emph{Epichloë}, which are vertically transmitted, as adaptive changes that improve the fitness of their hosts under increasing environmental stress.
This interpretation is in line with theory predicting that the positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors,donald2021context}.
We further found that trends in endophyte prevalence varied across the distribution of each species in association with changes in climate drivers, suggesting that the increases in endophyte prevalence are driven by context-dependent benefits to hosts that confer resilience under environmental change.
Taken together, this suggests an overall strengthening of host-symbiont mutualism over the last two centuries.

>>>>

\documentclass[11pt]{article}
%\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,round,sectionbib,sort]{natbib}   % omit 'round' option if you prefer square brackets
\let\cite\citep

@@

@@

	\section*{Abstract}
\linelabel{R2-C15-begin}Species' distributions and abundances are shifting in response to ongoing global climate change. 
Mutualistic microbial symbionts can provide hosts with protection from environmental stress that may \revise{promote resilience under} environmental change, however \revise{this change} may also disrupt species interactions and lead to declines in hosts and/or symbionts.
Symbionts preserved within natural history specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes) has changed over time in response to climate change, and how these changes vary across host species' \revise{distributions}.
Specifically, we \revise{examined} 2,346 herbarium specimens of three grass host species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, \emph{Elymus virginicus}) collected over the past two centuries (1824 -- 2019) for the presence or absence of \emph{Epichloë} symbiosis.
\revise{Analysis of an \jacob{approximate Bayesian spatially-varying coefficients model}{This is the first time INLA is mentioned. It seems to be a suggestion from the reviewer. However, I think most readers may not be familiar with it. I appreciate the precision regarding the type of Bayesian statistics (approximate), but I wonder if that might be too technical for the abstract. I would understand if it were included in the methods section.}\tom{}{I agree. I cut ' implemented in INLA'} revealed} that endophytes increased in prevalence over the last two centuries from ca. 25\% to ca. 75\% prevalence, on average, across three host species.
Changes in seasonal climate drivers were associated with increasing endophyte prevalence. 
Notably, increasing precipitation during the peak growing season for \emph{Agrostis} species and decreasing precipitation for \emph{E. virginicus} were associated with increasing endophyte prevalence.
Changes in the variability of precipitation and temperature during off-peak seasons were also important predictors of increasing endophyte prevalence. 
Our analysis performed favorably in an out-of-sample predictive test with \tom{contemporary survey data}{Since you drop numbers for the herbarium samples I would drop some numbers here too.}, a rare extra step in collections-based research.
We identified greater local-scale variability in endophyte prevalence in contemporary data compared to model predictions based on historic data, suggesting new directions that could improve predictive accuracy.
Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through \revise{fitness benefits to symbiotic hosts}. \linelabel{R2-C15-end}

@@

Natural history specimens, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial and temporal scales \citep{meineke2018unrealized}. 
Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines, primarily comprised of physical specimens of organisms along with information about the time and place of their collection. 
These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \linelabel{R2C19-begin}\tom{\revise{by examining reproductive structures}}{I don't really understand this revision or what prompted it. Phenology studies do not examine reproductive structures, they examine timing. I also don't think the sentence works grammatical with each list element having a 'by' clause.} \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \revise{through examination of pollen removal} \citep{pauw2011reconstruction, duan2019century}, and herbivory \revise{by documenting leaf damage} \citep{meineke2019herbarium} related to anthropogenic climate change. \linelabel{R2C19-end}
However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: microbial symbiosis.

Microbial symbionts are common to all macroscopic organisms and can have important effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts that can ameliorate environmental stress. 
For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner in ways that can cause the interaction to deteriorate \citep{aslan2013mutualism, fowler2024microbial}. 
Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the composition of their algal symbiont community \citep{meyer2022climate}.
How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions \citep{frederickson2017mutualisms}.
Previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualist partners \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.

Understanding of how microbial symbioses are affected by climate change is additionally complicated by spatial heterogeneity in the direction and magnitude of environmental change \cite{ipcc_2021}. 
Beneficial symbionts are likely able to shield their hosts from environmental stress in locations that experience a small degree of change, but symbionts in locations that experience changes of large magnitude may be pushed beyond their physiological limits \cite{webster2008temperature}.
Additionally, symbionts are often unevenly distributed across their host's distribution.
Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across their environments \cite{frade2008variation, rolshausen2018quantifying, fowler2023geographic}.
Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions for where to expect changes in the symbiosis in future climates.

\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grass species \citep{leuchtmann1992systematics}.
They are transmitted \tom{predominantly vertically}{Maybe two adverbs is awkward but I think we need to acknowledge the possibility of H transmission.} from maternal plants to offspring through seeds.
Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
Over time, endophytes that act as mutualists should rise in prevalence within a host population, particularly under environmental conditions that elicit protective benefits \citep{donald2021context}. 
\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
Previous survey studies of contemporary populations have documented large-scale spatial patterns in endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
We predicted that prevalence should track temporal changes in environmental drivers \linelabel{R2C21-begin}\revise{(i.e. drought)} \linelabel{R2C21-end} that elicit strong fitness benefits.
%For example, endophyte-mediated drought tolerance should lead endophyte prevalence to increase in regions where precipitation has declined over time.

Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of host species that harbor these symbionts \citep{white1985endophyte}, establishing that endophyte symbiosis could be identified in plant tissue from as early as 1851.
However, no subsequent studies, to our knowledge, have used the vast resources of biological collections to quantitatively assess spatio-temporal trends in endophyte prevalence and their environmental correlates. 
Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets. 
This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of this symbiosis to historical climate change. 
%Research using historical collections has clearly demonstrated other ecological signatures of a changing climate. 
However, the predictive ability derived from historical analyses is rarely tested against contemporary data \citep{lee2024phenological}. 
Critically evaluating whether insights from historical reconstruction are predictive of variation across contemporary populations is a crucial step for the field to move from reading signatures of the past to forecasting ecological dynamics into the future.

In this study, we assessed the long-term responses of \emph{Epichloë} endophyte symbiosis to climate change through the use of herbarium specimens of three North American host grass species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}).
We first addressed questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over the past two centuries? and (ii) How spatially variable are temporal trends in endophyte prevalence across eastern North America?
We then addressed how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between temporal trends in endophyte prevalence and associated changes in climate drivers?
\linelabel{R2C21-b-begin}We predicted that overall endophyte prevalence would increase over time in tandem with climate warming, and that localized hotspots of endophyte change would correspond spatially to hotspots of climate change. \linelabel{R2C21-b-end}
Finally, we evaluated \revise{(iv) how our model, built on data from historic specimens, performed in an out-of-sample test }using data on endophyte prevalence from contemporary surveys of host populations. 
To answer these questions we examined a total of $2,346$ historic specimens collected across eastern North America between 1824 and 2019, and evaluated model performance against contemporary surveys comprising 1,442 individuals from 63 populations surveyed between 2013 and 2020.

\section*{Methods}
        \subsection*{Focal species}
Our surveys focused on three native North American grasses: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus} that host \emph{Epichloë} symbionts. 
\linelabel{R2C23-begin}\revise{These cool-season grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States (Fig. \ref{fig:map}). Cool-season grasses grow during the cooler temperatures of spring and autumn due to their reliance on $C_3$ photosynthesis.}\linelabel{R2C23-end}
\emph{A. hyemalis} is a small short-lived perennial species that germinates in \tom{winter}{Winter, not spring. It's 'winter bentgrass'.} and typically flowers between March and July (most common collection month: May).
\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late summer and early autumn (most common collection month: September). 
\emph{A. perennans} is more sparsely distributed, tending to be found in shadier and moister habitats, while \emph{A. hyemalis} is commonly found in open and recently disturbed habitats. 
Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A. perennans} has a slighty more northern distribution, whereas \emph{A. hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed than the \emph{Agrostis} species. 
It begins flowering as early as March or April but continues throughout the summer (most common collection month: July).

\linelabel{R1C3-begin}
\revise{Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
The fungal symbionts primarily reproduce asexually and are passed from mother to offspring by vertical transmission through seeds. 
These traits contribute to highly specialized interactions between symbiont and host.
Some host species have been shown to partner with multiple symbiont species or strains, and in some cases multiple symbiont lineages can co-exist within a host population \cite{mc2014species}.
However, surveys have typically found limited \emph{Epichloë} genotypic diversity within host populations\cite{treindl2023genetic}. 
Across host populations, concentrations of biologically-active alkaloids and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.
In this analysis, we focus on the presence/absence of \emph{Epichloë} symbionts, and we discuss potential implications of symbiont genotypic diversity in the Discussion.}
\linelabel{R1C3-end}

		\subsection*{Herbarium surveys}
We visited nine herbaria between 2019 and 2022 (see \tom{Table}{In table, maybe replace -- with 0.} \ref{table:herbaria} for a summary of specimens included from each collection). 
With permission from herbarium staff, we acquired seed samples from $1135$ \emph{A. hyemalis} specimens collected between 1824 and 2019, $357$ \emph{A. perennans} specimens collected between 1863 and 2017, and $854$ \emph{E. virginicus} specimens collected between 1839 and 2019 (Fig. \ref{fig:map}, Fig. \ref{fig:temporal}A, Fig. \ref{fig:endo_status_map}).
%Our sampling plan was designed to minimize damage to these specimens.
We chose our focal species in part because they are commonly represented in herbarium collections and produce many seeds, meaning that small samples would not diminish the value of the specimens for future studies. 
We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
We excluded specimens for which information about the collection location and date were unavailable.

Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
Many specimens had digitized collection information readily available, but for those that did not, we transcribed information printed on the herbarium sheet. 
Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
For fifteen of the oldest specimens, only information at the state level was available, and so we used the state centroid.
\linelabel{R2C24-begin}\revise{The median pairwise distance between georeferenced coordinate points was 841 km.
The median longitudinal width of the bounding boxes generated to geocode municipality, county, or state centroids was 44.7 km. 
Among those specimens geo-referenced at the state level, the largest bounding box, spanning the state of Texas, was 1233 km wide.
The smallest bounding boxes were less than 1 km across for small municipalities (while this suggests high precision, we note that some specimens were collected in natural habitat nearby to small municipalities not encompassed by this bounding boxes).}\linelabel{R2C24-end}

\linelabel{R1C11-begin}
\revise{Our visits focused on herbaria with historic strengths in \emph{Poaceae} collections (e.g. Texas A\&M, Missouri Botanic Garden) and other herbaria in the Southern Great Plains region of the United States. 
While these nine herbaria garnered specimens that span the focal species' ranges, our dataset unevenly samples across the study \tom{region}{Reference Figure 1 here?}. 
Texas, Oklahoma, Louisiana, and Missouri are the most represented states.
Uneven sampling was most pronounced for \emph{A. perennans}, which has much of its range in the northeastern US. 
We explore the potential influence of spatial bias in sampling on our results through a simulation analysis (Appendix A - Supporting Methods).}\linelabel{R1C11-end}

After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae in each specimen using microscopy. 
We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue-lactic acid stain and squashed them under a microscope cover slip. 
We examined the squashed seeds for the presence of fungal hyphae at 200-400X magnification \cite{bacon2018stains}.
%In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded for that specimen.
On average we scored $4.7$ intact seeds per specimen of \emph{A. hyemalis}, $4.2$ seeds per specimen of \emph{A. perennans}, and $3.8$ seeds per specimen of \emph{E. virginicus}; we scored $10,342$ seeds in total.
\linelabel{R2C25-begin}Due to imperfect vertical transmission, \revise{the production of symbiont-free offspring from symbiotic hosts} \cite{afkhami2008symbiosis},\linelabel{R2C25-end} it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if \emph{Epichloë} hyphae were observed in none of its seeds. 
\linelabel{R2C26-begin}
To capture uncertainty in the endophyte \revise{identification} process, we recorded both a "liberal" and a "conservative" endophyte score for each plant specimen.  
\revise{When we confidently identified endophytes within a specimen's seeds, we assigned matching liberal and conservative scores.}
When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded \revise{ a positive identification} for the liberal score
and a negative \revise{identification for} the conservative score.
\tom{\revise{The liberal status assumed a potential endophyte identification was more likely to be endophyte-positive while the conservative status assumed that the potential endophyte identification was less likely to be endophyte-positive}. }{I don't think this sentence is necessary.}
$89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
The following analyses used the liberal status, \revise{however} repeating all analyses with the conservative status yielded qualitatively similar results (Fig. \ref{fig:conservative_scores_plot}).\linelabel{R2C26-end}

@@

\linelabel{R2C40-begin}Our examination of historic plant specimens revealed \revise{previously hidden} \linelabel{R2C40-end}shifts in microbial symbiosis over the last two centuries. 
For the three grass host species we examined, there have been strong increases in prevalence of\linelabel{R2C41-begin} \revise{\emph{Epichloë} endophyte symbiosis}.\linelabel{R2C41-end}
We interpret increases in prevalence of \emph{Epichloë}, which are vertically transmitted, as adaptive changes that improve the fitness of their hosts under increasing environmental stress.
This interpretation is in line with theory predicting that positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors,donald2021context}.
We further found that trends in endophyte prevalence varied across the distribution of each species in association with changes in climate drivers, suggesting that the increases in endophyte prevalence are driven by context-dependent benefits to hosts that confer resilience under environmental change.
Taken together, this suggests an overall strengthening of host-symbiont mutualism over the last two centuries.
